Unique solutions of contractions, CCS, and their HOL formalisation\thanksref{ATL1}

### 2020-07-31 ic-journal/sos.tex
Updated generated LaTeX sources

The CCS formalisation starts with type definitions for action,
relabeling and then the  processes. We use the type ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}''
for all labels (i.e. visible actions), where the type variable $\beta$ corresponds
to $\mathscr{L}$ (the set of names for labels) mentioned at the beginning
of Section~\ref{ss:ccs}. (Thus the
cardinality of ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' depends on its type variable:
when $\beta$ is finite or countable, ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' is countable.)
All labels are divided into input and output ones.
The type ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' is defined by HOL's \texttt{Datatype} package in the
following syntax:
\begin{lstlisting}
Datatype: Label = name 'b | coname 'b
End
\end{lstlisting}
Intuitively, ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' turns names in $\mathscr{L}$ into \emph{input and output labels}.
For instance, if the type $\beta$ is instantiated to \texttt{string}
(the type of all ASCII strings), then the HOL terms \HOLinline{\HOLConst{name} \HOLStringLit{a}} and
\HOLinline{\HOLConst{coname} \HOLStringLit{b}} denote the input label $a$ and output
label $\overline{b}$, respectively.
The type ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Action}}'' is the
union of all visible actions (input and output labels) and the invisible action $\tau$
(\texttt{tau}). For instace, the input \emph{action} $a$ and output
\emph{action} $\overline{b}$ of type ``\HOLinline{\HOLTyOp{string} \HOLTyOp{Action}}''
are denoted by \HOLinline{\HOLConst{label} \ensuremath{(}\HOLConst{name} \HOLStringLit{a}\ensuremath{)}} and \HOLinline{\HOLConst{label} \ensuremath{(}\HOLConst{coname} \HOLStringLit{b}\ensuremath{)}}, respectively. On the
other hand, \HOLinline{\HOLConst{label} \ensuremath{(}\HOLConst{name} \HOLNumLit{1}\ensuremath{)}} and \HOLinline{\HOLConst{label} \ensuremath{(}\HOLConst{coname} \HOLNumLit{2}\ensuremath{)}} could be actions of type ``\HOLinline{\HOLTyOp{num} \HOLTyOp{Action}}'', where
\HOLinline{\HOLTyOp{num}} is the type of natural numbers in HOL.

@@

In particular, the SOS rule \texttt{REC} (Recursion)
 says that if we substitute all occurrences of the variable $A$ in $P$ to
$(\recu A P)$ and the resulting process has a transition to $P'$
with an action $u$, then $(\recu A P)$ has the same
transition. Here ``\HOLinline{\ensuremath{[}\HOLConst{rec}\;\;\HOLFreeVar{X}\;\;\HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]}\;\;\HOLFreeVar{E}}'' is an abbreviation
for ``\HOLinline{\HOLConst{CCS_Subst}\;\;\HOLFreeVar{E}\;\;\ensuremath{(}\HOLConst{rec}\;\;\HOLFreeVar{X}\;\;\HOLFreeVar{E}\ensuremath{)}\;\;\HOLFreeVar{X}}'', where \HOLinline{\HOLConst{CCS_Subst}} is a
recursive function substituting all occurrences of a free variable with a CCS term.
For most CCS operators \HOLinline{\HOLConst{CCS_Subst}} just recursively goes into a deeper
level without changing anything, e.g.:
\begin{alltt}
\HOLTokenTurnstile{} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{(}\ensuremath{\HOLFreeVar{E}\sb{\mathrm{1}}} \HOLSymConst{\ensuremath{+}} \ensuremath{\HOLFreeVar{E}\sb{\mathrm{2}}}\ensuremath{)} \HOLSymConst{\ensuremath{=}} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{\HOLFreeVar{E}\sb{\mathrm{1}}} \HOLSymConst{\ensuremath{+}} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{\HOLFreeVar{E}\sb{\mathrm{2}}}\hfill{[CCS_Subst_sum]}
\end{alltt}
The only two insteresting cases are those for agent variables and recursion:
\begin{alltt}
\HOLTokenTurnstile{} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{(}\HOLConst{var} \HOLFreeVar{Y}\ensuremath{)} \HOLSymConst{\ensuremath{=}} \HOLKeyword{if} \HOLFreeVar{Y} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{X} \HOLKeyword{then} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{Y}\hfill{[CCS_Subst_var]}
\HOLTokenTurnstile{} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{Y} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\ensuremath{=}} \HOLKeyword{if} \HOLFreeVar{Y} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{X} \HOLKeyword{then} \HOLConst{rec} \HOLFreeVar{Y} \HOLFreeVar{E} \HOLKeyword{else} \HOLConst{rec} \HOLFreeVar{Y} \ensuremath{(}\ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}\hfill{[CCS_Subst_rec]}
\end{alltt}
Notice that variable substitutions only affect free
variables. For instance, the variable $Y$ in ``\HOLinline{\HOLConst{rec}\!\;\HOLFreeVar{Y}\!\;\HOLFreeVar{E}}'' is
bound and therefore the substitution ignores it.

>>>>

The CCS formalisation starts with type definitions for action,
relabeling and then the  processes. We use the type ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}''
for all labels (i.e. visible actions), where the type variable $\beta$ corresponds
to $\mathscr{L}$ (the set of names for labels) mentioned at the beginning
of Section~\ref{ss:ccs}. (Thus the
cardinality of ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' depends on its type variable:
when $\beta$ is finite or countable, ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' is countable.)
All labels are divided into input and output ones.
The type ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' is defined by HOL's \texttt{Datatype} package in the
following syntax:
\begin{lstlisting}
Datatype: Label = name 'b | coname 'b
End
\end{lstlisting}
Intuitively, ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}'' turns names in $\mathscr{L}$ into \emph{input and output labels}.
For instance, if the type $\beta$ is instantiated to \texttt{string}
(the type of all ASCII strings), then the HOL terms \HOLinline{\HOLConst{name} \HOLStringLit{a}} and
\HOLinline{\HOLConst{coname} \HOLStringLit{b}} denote the input label $a$ and output
label $\overline{b}$, respectively.
The type ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Action}}'' is the
union of all visible actions (input and output labels) and the invisible action $\tau$
(\texttt{tau}). For instance, the input \emph{action} $a$ and output % editorial
\emph{action} $\overline{b}$ of type ``\HOLinline{\HOLTyOp{string} \HOLTyOp{Action}}''
are denoted by \HOLinline{\HOLConst{label} \ensuremath{(}\HOLConst{name} \HOLStringLit{a}\ensuremath{)}} and \HOLinline{\HOLConst{label} \ensuremath{(}\HOLConst{coname} \HOLStringLit{b}\ensuremath{)}}, respectively. On the
other hand, \HOLinline{\HOLConst{label} \ensuremath{(}\HOLConst{name} \HOLNumLit{1}\ensuremath{)}} and \HOLinline{\HOLConst{label} \ensuremath{(}\HOLConst{coname} \HOLNumLit{2}\ensuremath{)}} could be actions of type ``\HOLinline{\HOLTyOp{num} \HOLTyOp{Action}}'', where
\HOLinline{\HOLTyOp{num}} is the type of natural numbers in HOL.

@@

In particular, the SOS rule \texttt{REC} (Recursion)
 says that if we substitute all occurrences of the variable $A$ in $P$ to
$(\recu A P)$ and the resulting process has a transition to $P'$
with an action $u$, then $(\recu A P)$ has the same
transition. Here ``\HOLinline{\ensuremath{[}\HOLConst{rec}\;\;\HOLFreeVar{X}\;\;\HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]}\;\;\HOLFreeVar{E}}'' is an abbreviation
for ``\HOLinline{\HOLConst{CCS_Subst}\;\;\HOLFreeVar{E}\;\;\ensuremath{(}\HOLConst{rec}\;\;\HOLFreeVar{X}\;\;\HOLFreeVar{E}\ensuremath{)}\;\;\HOLFreeVar{X}}'', where \HOLinline{\HOLConst{CCS_Subst}} is a
recursive function substituting all occurrences of a free variable with a CCS term.
For most CCS operators \HOLinline{\HOLConst{CCS_Subst}} just recursively goes into a deeper
level without changing anything, e.g.:
\begin{alltt}
\HOLTokenTurnstile{} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{(}\ensuremath{\HOLFreeVar{E}\sb{\mathrm{1}}} \HOLSymConst{\ensuremath{+}} \ensuremath{\HOLFreeVar{E}\sb{\mathrm{2}}}\ensuremath{)} \HOLSymConst{\ensuremath{=}} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{\HOLFreeVar{E}\sb{\mathrm{1}}} \HOLSymConst{\ensuremath{+}} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{\HOLFreeVar{E}\sb{\mathrm{2}}}\hfill{[CCS_Subst_sum]}
\end{alltt}
The only two interesting cases are those for agent variables and recursion: % editorial
\begin{alltt}
\HOLTokenTurnstile{} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{(}\HOLConst{var} \HOLFreeVar{Y}\ensuremath{)} \HOLSymConst{\ensuremath{=}} \HOLKeyword{if} \HOLFreeVar{Y} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{X} \HOLKeyword{then} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{Y}\hfill{[CCS_Subst_var]}
\HOLTokenTurnstile{} \ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{Y} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\ensuremath{=}} \HOLKeyword{if} \HOLFreeVar{Y} \HOLSymConst{\ensuremath{=}} \HOLFreeVar{X} \HOLKeyword{then} \HOLConst{rec} \HOLFreeVar{Y} \HOLFreeVar{E} \HOLKeyword{else} \HOLConst{rec} \HOLFreeVar{Y} \ensuremath{(}\ensuremath{[}\ensuremath{\HOLFreeVar{E}\sp{\prime}}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}\hfill{[CCS_Subst_rec]}
\end{alltt}
Notice that variable substitutions only affect free
variables. For instance, the variable $Y$ in ``\HOLinline{\HOLConst{rec}\!\;\HOLFreeVar{Y}\!\;\HOLFreeVar{E}}'' is
bound and therefore the substitution ignores it.
